Hilbert space approximate Bayesian Gaussian processes using Stan

### 2020-03-18 Paper/manuscript_v2/manuscript_07_01.tex
more edits across the paper

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters $f$ and $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes trivially to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*}

@@

If $p(y_n|f(\bm{x}_n),\phi)=N(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GP and $O(nm^2)$ for sparse GP). If $p(y_n|f(\bm{x}_n),\phi)=N(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} need to be used. In this paper we focus on use of MCMC for integrating over the joint
posterior. MCMC is not usually the fastest approach, but allows
accurate inference for general models in probabilistic programming
setting. We consider the computational costs of GPs specifically from
this point of view.

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\sigma^2 \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\nu}K_{\nu}\left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell$ and $\sigma$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\sigma^2 \exp\bigg(-\frac{1}{2} \frac{r^2}{\ell^2}\bigg),  \\
k_{\frac{3}{2}}(r)&=\sigma^2\bigg(1+\frac{\sqrt{3}r}{\ell}\bigg) \exp\bigg(-\frac{\sqrt{3}r}{\ell}\bigg). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form:
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\sigma^2 \exp\bigg(-\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\bigg),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\sigma^2 \Bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg)\exp\Bigg(- \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. The use of a multidimensional length-scale basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \sigma^2 \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\omega^2 \right)^{(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\sigma$ are the lengthscale and magnitud (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form:
%
\begin{align}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form:
%
\begin{align*}
S_{\infty}(\omega)&= \sigma^2 \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \mathrm{exp}\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \right),   \\
S_{\frac{3}{2}}(\omega)&= \sigma^2 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \right)^{-3}.
\end{align*}

@@

The approximate GP method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}. Our starting point for presenting the method is the main result obtained by \cite{solin2018hilbert} of the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation have been briefly presented in the Appendix \ref{ch5_app_approx_covfun} of this paper.

@@

Ultimately, these recommendations lie on the relationships among the number of basis functions, the boundary factor and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we built these relationships for the square exponential covariance function and Matern ($\nu$=3/2) covariance function in the present section, and for the periodic squared exponential covariance function in Appendix \ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different, in function of mainly the smoothness or wigglyness of the kernel effects.

@@

\begin{itemize}
\item As $\ell$ increases, $c$ and $m$ required for a close enough approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve a close approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a close approximation is never going to be achieved regardless of $m$. This fact can be appreciated in the Figure as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$).
\end{itemize}

As stated above, Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c}. If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale provided by Figure \ref{fig6_lscale_vs_J_vs_c}, repeat the process while increasing $m$ or decreasing $c$. Note that, as commented before, $c$ can not be decreased as much as desired because it is restricted to the lengthscale.

@@

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} were build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which can not be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wiggliness in every dimensions. For instance, in the two-dimensional case it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} to check the accuracy by analyze individually the different dimensions of a multidimensional GP model.

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the regular GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both regular GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). Notice that the dashed red line of the regular GP can hardly be seen in the plots because it is under the blue lines.}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

@@

\section{Univariate case studies}\label{ch5_sec_univariate_cases}

\subsection{Simulated data}\label{ch5_sec_univariate_simu}

This example consists of a simulated dataset with $n=250$ ($i=1,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise $\sigma=0.2$ was added to the GP draws.

@@

In the HSGP model, the latent function values $f(x)$ are approximated as in equation (\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in equation (\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively.

In order to do model comparison, in addition to the regular GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2015package}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

Figure \ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The computational time is represented in the y-axis of the figure, which is in a logarithmic scale. The HSGP model is on average roughly 400 times faster than the regular GP, in the particular case of applying over this dataset.

The Stan model codes for the exact GP, the approximate GP and the spline models of this case study can be found through the following link:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-I_1D-Simulated-data
\end{lstlisting}

@@

The Stan model code for the approximate GP model of this case study can be found in:
%
\begin{lstlisting}[breaklines]
https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-II_Birthday-data
\end{lstlisting}

@@

\section{Multivariate case studies}\label{ch5_sec_multivariate_cases}

>>>>

In this work, GPs will take the role of a prior distribution over function spaces for non-parametric latent functions in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \bm{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is modelled conditionally as $p(y_n|f(\bm{x}_n),\phi)$, where $p$ is some parametric distribution with parameters $f$ and $\phi$, and $f$ is an unknown function with GP prior, which depends on an input $\bm{x}_n\in {\rm I\!R}^D$. This generalizes readily to more complex models depending on several unknown functions, for example such as $p(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or multilevel models. Our goal is to obtain posterior distribution for the value of the function $\tilde{f}=f(\tilde{\bm{x}})$  evaluated at a new input point $\tilde{\bm{x}}$.
% That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.\\

@@

The mean and covariance functions completely characterize the GP prior, and control the a priori behavior of the function $f$. Let $\bm{f}=\left\lbrace f(\bm{x}_n) \right\rbrace_{n=1}^N$, then the resulting prior distribution for $\bm{f}$ is a multivariate Gaussian distribution
$\bm{f} \sim \mathcal{N}(\bm{\mu}, \bm{K})$,
 where $\bm{\mu} = \left\lbrace \mu(\bm{x}_n) \right\rbrace_{n=1}^N$ is the mean and $\bm{K}$ the covariance matrix, where $K_{i,j}=k(\bm{x}_i,\bm{x}_j)$. In the following, we focus on zero-mean Gaussian processes, that is set $\mu(\bm{x}) = 0$. The covariance function $k(\bm{x}, \bm{x}')$ might depend on a set of hyperparameters, $\bm{\theta}$, but we will not write this dependency explicitly to ease the notation.
The joint distribution of $\bm{f}$ and a new $\tilde{f}$ is also a multivariate Gaussian as,
%
\begin{align*}
p(\bm{f}, \tilde{f})=\mathcal{N} \left( \left[ \begin{array}{cc}
\bm{f} \\ 
f^*
\end{array} \right] \,\middle|\, \bm{0},\left[ \begin{array}{cc}
\bm{K}_{\bm{f},\bm{f}} & \bm{k}_{\bm{f},\tilde{f}} \\ 
\bm{k}_{\tilde{f},\bm{f}} & k_{\tilde{f},\tilde{f}}
\end{array} \right] \right),
\end{align*}

@@

If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),\sigma)$ then $\bm{f}$ can be integrated out analytically (with a computational cost of $O(n^3)$ for exact GPs and $O(nm^2)$ for sparse GPs). If $p(y_n|f(\bm{x}_n),\phi)=\mathcal{N}(y_n|f(\bm{x}_n),g(\bm{x}_n))$ or $p(y_n|f(\bm{x}_n),\phi)$ is non-Gaussian, the marginalization does not have closed form solution. Furthermore, if a prior distribution is imposed on $\phi$ and $\bm{\theta}$ to form a joint posterior for $\phi$, $\bm{\theta}$ and $\bm{f}$, approximate inference such as Markov chain Monte Carlo (MCMC) \citep{brooks_2011}, Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational Bayes methods \citep{gibbs2000variational,csato2000efficient} need to be used. In this paper, we focus on the use of MCMC for integrating over the joint
posterior. MCMC is not usually the fastest approach, but allows
accurate inference and uncertainty estimates for general models in probabilistic programming
settings. We consider the computational costs of GPs specifically from
this point of view.

@@

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\nu}K_{\nu}\left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}

\noindent where $\nu$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp\bigg(-\frac{1}{2} \frac{r^2}{\ell^2}\bigg),  \\
k_{\frac{3}{2}}(r)&=\alpha\bigg(1+\frac{\sqrt{3}r}{\ell}\bigg) \exp\bigg(-\frac{\sqrt{3}r}{\ell}\bigg). 
\end{align*}

\noindent The former is commonly known as squared exponential (exponentiated quadratic) covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form:
%
\begin{align*}
k_{\infty}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \exp\bigg(-\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\bigg),\\
k_{\frac{3}{2}}(|\bm{x}-\bm{x}'|_{L2})&=\alpha \Bigg(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg)\exp\Bigg(- \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\Bigg). \\
\end{align*}

\noindent Notice that the previous expressions have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. The use of a multidimensional length-scale basically turns the isotropic covariance function into a non-isotropic one. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure (\textit{Bochner's theorem}, see, e.g. \cite{akhiezer1993theory}). If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions is given by
%
\begin{align*}
S_{\nu}(\omega)&= \alpha \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)l^{2\nu}}\left(\frac{2\nu}{l^2}+4\pi^2\omega^2 \right)^{(\nu+D/2)}
\end{align*}

\noindent in $D$ dimensions, where variable $\omega\in {\rm I\!R}$ is a distance in the frequency domain, and $\ell$ and $\alpha$ are the lengthscale and magnitude (marginal variance), respectively, of the kernel. The particular cases where $\nu=\infty$ and $\nu=3/2$ take the form:
%
\begin{align}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^D  \ell^D  \exp\left(-0.5 \ell^2 \omega^2\right), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\omega)&= \alpha \frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2})\sqrt{3}^3}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\omega^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}

\noindent Particularizing to an input dimension $D=3$ and Euclidean distance $\omega=\sqrt{\sum_{i=1}^{3}s_i^2}$, and considering a multidimensional lengthscale $\bm{\ell} \in {\rm I\!R}^{3}$, the spectral densities written above take the form:
%
\begin{align*}
S_{\infty}(\omega)&= \alpha \sqrt{2\pi}^{3}  \prod_{i=1}^{3} \ell_i  \mathrm{exp}\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 s_i^2 \right),   \\
S_{\frac{3}{2}}(\omega)&= \alpha 32\pi\sqrt{3}^3\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 s_i^2 \right)^{-3}.
\end{align*}

@@

The approximate GP method, developed by \cite{solin2018hilbert} and implemented in this paper, is based on considering the covariance operator of a homogeneous (stationary) covariance function as a pseudo-differential operator constructed as a series of Laplace operators. Then, the pseudo-differential operator is approximated with Hilbert space methods on a compact subset $\Omega \subset {\rm I\!R}^D$ subject to some boundary condition. For brevity, we will refer to these approximate Gaussian processes as HSGPs. Below, we will present the main results around HSGPs relevant for practical application. More details and mathematical proofs are provided in \cite{solin2018hilbert}. Our starting point for presenting the method is the main result obtained by \cite{solin2018hilbert} of the definition of the covariance function as a series expansion of eigenvalues and eigenfunctions of the Laplacian operator. The mathematical details of this approximation are briefly presented in Appendix \ref{ch5_app_approx_covfun}.

@@

Ultimately, these recommendations are based on the relationships among the number of basis functions, the boundary factor and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we built these relationships for the square exponential covariance function and Matern ($\nu$=3/2) covariance function in the present section, and for the periodic squared exponential covariance function in Appendix \ref{ch5_sec_periodic}. For other kernels, the relationships will be slightly different, in function of mainly the smoothness or wigglyness of the kernel effects.

@@

\begin{itemize}
\item As $\ell$ increases, $c$ and $m$ required for a close enough approximation decrease.
\item The lower $c$, the smaller $m$ can and $\ell$ must be to achieve a close approximation.
\item For a given $\ell$ there exist a minimum $c$ under which a close approximation is never going to be achieved regardless of $m$. This fact can be seen in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} as the contour lines which represent $c$ have an end in function of $\ell$ (Valid $c$ are restricted in function of $\ell$).
\end{itemize}

As stated above, Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} provide the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c} or \ref{fig6_lscale_vs_J_vs_c_Matern} (depending on which kernel is used). If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, $\hat{\ell}$ does not exceed recommended minimum lengthscale, the approximation may be inaccurate and $m$ should be increased or $c$ decreased. We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale, repeat the process while increasing $m$ or decreasing $c$.
As mentioned earlier, $c$ cannot be decreased as much as desired because it is restricted by the lengthscale, so increasing $m$ may usually the preferred approach.

@@

Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} were build for a GP with a unidimensional covariance function, which result in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which is more difficult to be graphically represented. More precisely, in the multi-dimensional case, whether the approximation is close enough might depend only on the ratio between wiggliness in every dimensions. For instance, in the two-dimensional case, it would depend on the ratio between $\ell_1$ and $\ell_2$ and could be graphically represented. Future research will focus on building useful graphs or analytical models that provide these relations in multi-dimensional cases. However, as an approximation, we can use the unidimensional GP conclusions in Figures \ref{fig6_lscale_vs_J_vs_c} and \ref{fig6_lscale_vs_J_vs_c_Matern} to check the accuracy by analyze individually the different dimensions of a multidimensional GP model.

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the regular GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.
For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.42, trim = 10mm 100mm 5mm 30mm, clip]{ch5_fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 25.5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.31, trim = 1mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.31, trim = 21mm 5mm 10mm 19mm, clip]{ch5_fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both regular GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($\ell$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{true function}). Notice that the dashed red line of the regular GP can hardly be seen in the plots because it is almost perfectly overlayed by the blue lines.}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

@@

%\section{Univariate case studies}\label{ch5_sec_univariate_cases}
\section{Case studies}\label{ch5_sec_cases}

In this section, we will present several simulated and real case studies in which we apply the developed HSGP models. More case studies are presented in the online supplemental materials.

\subsection{Simulated data for a univariate function}\label{ch5_sec_univariate_simu}

This example consists of a simulated dataset with $n=250$ ($i=1,\dots,n$) single draws from a Gaussian process prior with a Mat{\'e}rn($\nu$=3/2) covariance function and hyperparameters marginal variance $\alpha=1$ and lengthscale $\ell=0.15$, with corresponding inputs values $\bm{x}=(x_1\dots,x_n)$ with $x_i \in [-1,1] \subset {\rm I\!R}$. To form the final noisy dataset $\bm{y}$, Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws.

@@

In the HSGP model, the latent function values $f(x)$ are approximated as in equation (\ref{eq_approxf}), with the Mat{\'e}rn($\nu$=3/2) spectral density $S$ as in equation (\ref{eq_specdens_32}), and eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ as in equations (\ref{eq_eigenvalue}) and (\ref{eq_eigenfunction}), respectively. 
In order to do model comparison, in addition to the regular GP model and HSGP model, a spline-based model is also fitted using the thin plate regression spline approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2015package}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

Figure \ref{ch5_fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the spline model. The HSGP model is on average roughly 400 times faster than the regular GP, in this particular application.

The Stan model code for the exact GP, the approximate GP and the spline models of this case study can be found online at \url{https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-I_1D-Simulated-data}.
%

@@

The Stan model code for the approximate GP model of this case study can be found at \url{https://github.com/gabriuma/Doctoral_thesis/tree/master/Case-study-II_Birthday-data}.

@@

%\section{Multivariate case studies}\label{ch5_sec_multivariate_cases}
